---
title: "Tooth Growth Analysis"
author: "Jason T. Brown"
date: "March, 2015"
output: pdf_document
---

## Overview
The effect of two dietary supplements on guinea pig tooth growth is studied. Doses of vitamin C over 0.5 mg encourage tooth growth whether delivered by orange juice or ascorbic acid. This document is a literate statistical program written in R.
 
## Exploratory Data Analysis
I first retrieve the data from R's built in datasets library, and print a summary of the variables. Using `help(ToothGrowth)` reveals the variables are the length of guinea pig teeth in unspecified units, the supplement delivery method, either orange juice ("OJ") or ascorbic acid ("VC"), and dose in milligrams.

```{r getdata, results='asis'}
library(xtable)
data(ToothGrowth)
print(xtable(summary(ToothGrowth)), type="latex", comment=FALSE)
```

The initial summary indicates the data set is small, 60 observations, and has no missing values. I next check the number of subjects for each supplement and dose.  

```{r balanced, results='asis'}
print(xtable(table(ToothGrowth[, 2:3])), type="latex", comment=FALSE)
```

As a final exploratory activity, I plot the tooth length versus the dose and supplement. Note the use of the [violin plot](http://en.wikipedia.org/wiki/Violin_plot). This reveals a general upward trend in length with the dose. It appears the orange juice may be more effective than ascorbic acid at lower doses.

```{r plotExplore}
library(ggplot2)
p <- ggplot(ToothGrowth, aes(x = supp, y = len, group = dose))
p + facet_grid(. ~ dose) + geom_violin(aes(group = supp)) +  geom_point(aes(col = supp)) +
 labs(title = "ToothGrowth Summary", x = "supp within dose")
```

## Inference Tests

###Dose
I test the effect of dose and supplement using two inference tests. There are many inference tests which could be conducted even on this simple dataset. However I limit these intentionally to avoid fishing for significant results. The first test is whether the mean of tooth length is different at different dose levels. I use a two group *t*-test on the group of 0.5 mg doses versus larger doses.  The significance level is 95%. The null hypothesis is that the means are equal; the alternative is that the smaller doses have a smaller mean.

```{r infTestDose}
t.test(subset(ToothGrowth, dose<=0.5, len), subset(ToothGrowth, dose>.5, len),
       alternative = "less", paired = FALSE, var.equal=FALSE)
```

I conclude to reject the null hypothesis in favor of the alternative that the small doses have a smaller mean. Observe the confidence interval indicates that the means differ by at least 10.13, at the 95% significance level.

###Supplement
I next test, within the large doses, for the effect of the supplement. Formally, I will test the difference of the group means for the two supplments. The null hypothesis is that the means are equal and the alternative is that they are not equal. I again test at 95% significance level.

```{r infTestSupp}
t.test(subset(ToothGrowth, supp=="VC" & dose>0.5, len), 
       subset(ToothGrowth, supp=="OJ" & dose>0.5, len),
       alternative = "two.sided", paired = FALSE, var.equal=TRUE)
```

I fail to rejec the null hypothesis at 95% significance level. I conclude that the supplement means are the same at doses of 1 or 2 mg.

### Assumptions
In the inference test on doses, I did not assume equal variances. This requires further  assumptions about the applicability of the *t*-statistic. It also assumes that each population variance is approximately equal to the sample variance. In the supplement inference test I assumed the variances were equal for either supplment. 

**Bit more about t test generally**

## Conclusions
The conclusions for this study depend on whether tooth growth is considered desirable or not. In either case, a higher dose of either orange juice or ascorbic acid is more effective. It is not clear which supplement is superior at doses of 1 or 2 mg. This study cannot predict the effect of doses higher than 2 mg.

Additional research is warranted if an economical plan to encourage tooth growth is to be formulated. Other analysis techniques could be applied to the existing data. Those results could augment the conclusions of this study to formulate subsequent experimental designs.

