---
title: "Transmission Choice and Fuel Economy"
author: "Jason T. Brown"
date: "March 2015"
output: pdf_document
---
```{r libs,echo=FALSE}
library(xtable)
library(ggplot2)
#NTE 2 pages body ---
```

# Executive Summary

In this study, regression models are used to determine the effect of automatic or manual transmission on fuel efficiency. Generally, cars with automatic transmissions are estimated to achieve 3.1 to 7.5 mpg worse economy. The analysis also showed that other design factors such as horsepower and weight are very influential. This is a literate statistical programming document using R code embedded throughout.[^1]

# Data Exploration
```{r getData}
data(mtcars)
mtcars$tran <- factor(mtcars$am, levels = c(1,0), labels = c("manual", "auto"))
```

The data are built in to R. I plot them by quarter-mile time and engine displacement, shown in the appendix. These two variables provide a clean break between automatic and manual cars. 

# Regression Modeling
I answer the study questions using regression modeling. The approach is to describe the fuel economy as an outcome of characteristics of the car. I am interested in models that include the transmission type, as these will indicate the direction and size of the effect on fuel economy.

## Model Selection
The model selection process began by fitting a simple model with only transmission predicting economy. From here, additional candidate variables were added one at a time, with the interaction term. The best model found by this approach uses only weight and horsepower with interaction (adjusted $R^2 = 0.87$). This result confirms our intuition about fuel economy generally. The best model of fuel economy found which also includes transmission is: $mpg = 31.9 - 0.059 \cdot hp - 5.28 \cdot auto$. 

```{r hpModel}
hp <- lm(mpg ~ tran + hp, mtcars)
summary(hp)
```

## Model Diagnostics
The diagnostic performance of the model is reasonably good. There is nothing to suggest a data error. None of the points seem to have excessive influence on the model. Diagnostic plots are shown in the appendix. The observed mpg value for two cars is above the maximum predicted by the model. These are shown below and are also highlighted in the diagnostic plots.  
```{r overIntercept,results='asis'}
print(xtable(mtcars[mtcars$mpg >= hp$coefficients[1], c(1,4,12)]), type="latex", comment=F)
```

## Interpretation
The strict interpretation of the model is that automatic transmission is worse for fuel economy. Compared to manual, automatic means 5.28 less miles per gallon on average for cars with the same horsepower. The model also indicates that each additional horsepower rated translates into 0.059 less mpg, or 5.9 less miles per gallon per 100 hp. The 95% confidence intervals for mpg change on both transmission and 100 hp are computed below.  The regression model, with confidence interval, is plotted in the appendix. 

```{r tranautoCI, results='asis'} 
co <- summary(hp)$coefficients
tranauto <- co["tranauto","Estimate"] + c(-1, 0, 1) * qt(.975, df = hp$df) * 
    co["tranauto","Std. Error"]
names(tranauto) <- c("95% CI lower", "estimate", "95% CI upper")
hpCI <- (co["hp","Estimate"] + c(-1, 0, 1) * qt(.975, df = hp$df) * 
             co["hp","Std. Error"]) *100
print(xtable(rbind(tranauto,hpCI)), type="latex",comment=F)
```


[^1]: This open source analysis was prepared by @vpipkt using [Rmarkdown](http://rmarkdown.rstudio.com). Contribute on [github](https://github.com/vpipkt/datasciencecoursera/tree/master/Regression/project).


\pagebreak

# Appendix

## Exploratory analysis
```{r explore}
d <- ggplot(mtcars, aes(x = disp, size = mpg, y = qsec, col=tran))
d + geom_point() + xlab("Engine displacement, cu. in.") +  ylab("1/4 mile time, sec")
```

## Diagnostic plot
```{r diagPlots,fig.height=6}
par(mfrow = c(2,2))
plot(hp)
```

## Regression model with confidence intervals
```{r modelWithCI}
# model predictions with confidence interval
pred <- predict(hp, interval="confidence")
predframe <- with(mtcars,
                  data.frame(hp, tran, mpg = pred[, "fit"], 
                             lwr = pred[, "lwr"], upr = pred[, "upr"]))
# model coefficients in data frame, for geom_abline
coeffs <- data.frame(tran=factor(c("auto", "manual")), 
                     a = c(sum(hp$coefficients[1:2]),hp$coefficients[1]), 
                     b = rep(hp$coefficients[3],2))

modelPlot <- ggplot(mtcars, aes(x = hp, y = mpg))
modelPlot + geom_ribbon(data = predframe, aes(ymin = lwr, ymax = upr, group = tran), 
                        fill = "grey70", alpha = 0.5) +
    geom_abline(data = coeffs, aes(intercept = a, slope = b, color = tran)) + 
    geom_point(aes(col = tran), size = 3) +
    labs(title = "Fitted model mpg ~ tran + hp")
```

